---
  title: "Extracting and Cleaning the bulk-FRET Binding Data"
author: "Brady Johnston"
date: "2021-10-08"
output:
  workflowr::wflow_html:
  code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
```

# Getting the Data

```{r}
fl <- list.files("~/Dropbox/BondLab/Data/Typhoon Images/190830/imagejprocessed/", pattern = ".csv", 
           full.names = TRUE)
# fl[1]
lapply(fl, function(x) {
  df <- readr::read_csv(x, col_types = cols(), progress = TRUE)
  df$file <- x
  df
}) %>% 
  do.call(rbind, .) -> df

df <- df %>% 
  select(Min, Max, Row, Column, Absorbance, file) %>% 
  mutate(file = str_extract(file, "(?<=\\/\\/).+(?=.tif)")) %>% 
  mutate(sample = str_extract(file, "\\d{2}_\\d{2}"), 
         channel = str_extract(file, "(?<=-).+$")) %>% 
   mutate(channel = if_else(str_detect(channel, "FRET"), "FRET", channel))
```

## Peak at the plate data.

We can have a quick look at all of the plate data, to remember how the rows and columns were arranged.

```{r}
df %>%
  filter(channel == "FRET", sample == "01_02") %>%
  mutate(Row = factor(Row, levels = c(LETTERS[8:1]))) %>%
  ggplot(
    aes(
      x      = Column,
      colour = Absorbance,
      y      = Row
      )) +
  geom_point(size = 12) +
  scale_colour_viridis_c(option = "B",
                         limits = c(3.2, NA),
                         direction = 1) +
  theme_light() +
  scale_x_continuous(position = "top", breaks = c(1:12)) +
  theme(aspect.ratio = 8 / 12)
```

We can see that the increase intensity goes across the columns, so the RNA were in the rows A-C and D-F respectively, with G being the protein-only sample. Now we can calculated the FRET from the *transfer* and *donor* channels.

```{r}
df <- df %>% 
  mutate(
    rna = case_when(
      Row %in% c(c("A", "B", "C")) ~ str_extract(sample, "\\d{2}(?=_)"),
      Row %in% c(c("D", "E", "F")) ~ str_extract(sample, "(?<=_)\\d{2}"), 
      Row %in% "G" & Column %in% c(1:3) ~ "negative", 
      TRUE ~ "empty"
    ), 
    conc = 1e3 / 2 ^ (Column - 1)
  )
df
```

The RNA and the concentrations have been assigned, we can now also `pivot_wider()` to have a separate column for each channel.

```{r pivot-wider}
df <- df %>% 
  pivot_wider(
    names_from = channel, 
    values_from = Absorbance, 
    id_cols = c(rna, sample, Row, Column, conc)
  ) %>% 
  unnest(Cy3, Cy5, FRET) %>% 
  rename(row = Row, col = Column, don = Cy3, acc = Cy5, trans = FRET) %>% 
  mutate(FRET = trans / (don + trans))
```

Hmm that doesn't seem right, why aren't there values in the Cy5 and the FRET columns? Seems that it is incorrectly grouping things separately.

```{r}
df %>% 
  filter(as.numeric(rna) %in% 1:13) %>% 
  ggplot(aes(conc, FRET, colour = rna)) + 
  geom_point() + 
  facet_wrap(~rna, scales = "free_y") + 
  scale_x_log10()

fit_df <- df %>% 
  mutate(id = paste(sample, rna, sep = ":")) %>% 
  biochemr::b_binding(conc, FRET, id)

fit_df %>% 
  filter(!str_detect(id, "empty|negative")) %>%
  biochemr::b_plot(colour = id) + 
  facet_wrap(~id, scales = "free_y") + 
  scale_x_log10()

source("~/git/thesis/figures/chap4/bulk-FRET/rna-binding.R")

df <- rename(df, rna = sample)

df <- df %>% filter(!(rna %in% c("05", "06")))

trimmed_fit <- fit_df %>% filter(str_detect(id, ":01|:02|:05|:06")) %>% 
  rename(rna = id) %>% 
  mutate(rna = str_extract(rna, "(?<=:)\\d{2}")) %>% 
  .[2:5, ]

cdf <- rbind(trimmed_fit, df)

cdf
```

Plot out the values.

```{r}
cdf %>% 
  arrange(rna) %>% 
  
  biochemr::b_plot()
```
