---
title: "Looking at the Traces from Bishnu"
author: "Brady Johnston"
date: "2021-09-16"
output: 
    workflowr::wflow_html:
      code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
```


## Introduction

Making a fret plotting function to check the data.

```{r}
fl <- "~/Dropbox/BondLab/Data/smFRET/bishnu_brady_shared/1 nM/"

fretplot_function <- function(x) {
  df <- read_table(x, 
                   col_types = cols(),
             col_names = c("frame", "don", "acc", "fret", "state"))
  
  df <- df %>% 
    pivot_longer(cols = !contains("frame"))
  
  signal_plot <- df %>% 
    filter(name %in% c("don", "acc")) %>% 
    ggplot(aes(frame, value, colour = name)) + 
    geom_line() + 
    theme_classic() + 
    labs(y = "Fluoroscent signal") +
    guides(colour = "none") +
    theme(axis.text.x = element_blank(), 
          axis.ticks.x = element_blank(), 
          axis.title.x = element_blank())
  
  fret_plot <- df %>% 
    filter(!(name %in% c("don", "acc"))) %>% 
    ggplot(aes(frame, value, linetype = name)) + 
    geom_line() + 
    scale_y_continuous(breaks = 0:5*0.2) +
    coord_cartesian(ylim = c(0,1)) +
    labs(y = "FRET", 
         x = "Frame") +
    theme_classic() + 
    scale_colour_manual(values = c("black", "midnightblue"))
  
  
  
  final_plot <- patchwork::wrap_plots(signal_plot, fret_plot, ncol = 1)
  final_plot  
}


```

Plot one of the .dat files.

```{r}
list.files(fl, ".dat", full.names = TRUE)[3] %>% fretplot_function
```

Looks good, lets have a look what is in the merged files.

```{r}
list.files(here::here(fl, "merged"), full.names = TRUE) %>% 
  .[1] %>% 
  fretplot_function()
```

Seems that they are just more of the same? Unsure what the difference between them are.

```{r}
list.files(here::here(fl, "merged"), full.names = TRUE) %>% 
  .[100] %>% 
  fretplot_function()
```

Let's try remaking some of the plots that bishnu sent through.

```{r}
files <- list.files(here::here(fl, "merged"), full.names = TRUE)

read_all_frames <- function(files) {

  lapply(seq_along(files), function(x) {
    df <- read_table(files[x], 
                     col_types = cols(),
                     col_names = c("frame", "don", "acc", "fret", "state"))
    df$mol <- x
    
    df
  }) %>% 
    do.call(rbind, .) -> combined_mols
  
  combined_mols
}

df <- read_all_frames(list.files(fl, full.names = TRUE, pattern = ".dat"))

df

```

Looking good, now to try and make the plots.
```{r}
time_density_plot <- function(data) {
  data %>% 
    filter(fret > 0, fret < 1) %>% 
    mutate(frame = frame / 5) %>% 
    ggplot(aes(frame, fret)) + 
    geom_point(alpha = 0.1, colour = "gray30") +
    stat_density_2d(aes(fill = ..ndensity.., alpha = ..ndensity..),
                    geom = "raster",
                    n = 200,
                    contour = FALSE) +
    geom_density_2d(bins = 10, colour = "white", contour_var = "ndensity") +
    scale_fill_viridis_c(option = "D",
                         direction = 1,
                         limits = c(0.002, NA),
                         # trans = "log10",
                         na.value = NA) +
    scale_alpha_continuous(range = c(0.01, 0.8)) +
    coord_cartesian(ylim = c(0,1), xlim = c(0,100)) + 
    guides(alpha = "none", 
           fill = "none") +
    # theme_light() +
    theme_linedraw() +
    labs(x = "Time (s)", y = "FRET") +
    scale_x_continuous(expand = c(0,0)) + 
    scale_y_continuous(expand = c(0,0))
  
}
```

```{r}
plot_1nm <- df %>% 
  time_density_plot() + 
  labs(title = "1 nM Target")
```

```{r}
files_01 <-
  list.files(
    "~/Dropbox/BondLab/Data/smFRET/bishnu_brady_shared/100 pM_RNA1/",
    pattern = ".dat",
    full.names = TRUE
  )

df_01 <- read_all_frames(files_01)

plot_1pm <- df_01 %>% 
  time_density_plot() + 
  labs(title = "100 pM Target")
```


```{r , fig.height=4, fig.cap="Time-density plots for subpopulations of molecules."}
patchwork::wrap_plots(plot_1nm, plot_1pm)
```

It seems that the plots from Bishnu's data can be replicated. Now I just need to do so with the the other traces to get a more complete picture.


## Plotting the Static Experiments

Below is the FRET data from the static experiments, where the molecules were incubated with their RNA target for 15 minutes and then the movies were taken.

```{r static-plot}
fl <- list.files(
  "~/Dropbox/BondLab/Data/smFRET/bishnu_brady_shared/Bishnu/RNA1_target/", 
  full.names = TRUE, 
  pattern = ".dat"
  )

read_fun <- function(file) {
  some_list <- pbapply::pblapply(seq_along(file), function(x) {
    df <- readr::read_table(file[x], 
                            col_types = readr::cols(), 
                            col_names = c("frame", "don", "acc", "fret", "state"))
    df$id <- x
    df
  })
  
  do.call(rbind, some_list)
}

df <- read_fun(fl)

df %>% 
  time_density_plot()
```

## Plotting the Live-Flow experiments.

This is the live-flow experiments, where the movies movies were recored, as the RNA was injected into the flow-cell to capture the live binding events of the proteins.
```{r}
#| fig.height = 8,
#| fig.retina = 2

fl2 <- list.files(
  "~/Dropbox/BondLab/Data/smFRET/bishnu_brady_shared/Bishnu/RNA1_target/liveflow_RNA1/",
  full.names = TRUE
  )

live_rna01 <- read_fun(fl2)

live_rna01 %>% 
  time_density_plot()
```


# Applying to all of the RNA

Need to apply this approach to all of the RNA to replicated all of their plots. To do that we'll need to read in each of the sets of plots.
```{r read-all}
paths <- c(
  "data/smfret-data/RNA1_target",
  "data/smfret-data/RNA2",
  "data/smfret-data/RNA3",
  "data/smfret-data/RNA4",
  "data/smfret-data/RNA5",
  "data/smfret-data/RNA6",
  "data/smfret-data/RNA7",
  "data/smfret-data/RNA8",
  "data/smfret-data/RNA9",
  "data/smfret-data/RNA10",
  "data/smfret-data/RNA11",
  "data/smfret-data/RNA12",
  "data/smfret-data/RNA13"
)
list.files(paths[8], ".dat")

some_list <- pbapply::pblapply(seq_along(paths), \(x) {
  df <- read_fun(list.files(paths[x], ".dat", full.names = TRUE))
  df$rna <- x
  df
})

df <- do.call(rbind, some_list[c(1:7, 9:13)])

df %>% 
  time_density_plot() + 
  facet_wrap(~rna)
```

Now we have read all of it in, and are able to get time-density plots, lets try to replicate the transition-density plots (TDP). To achieve this, need to achieve this I need to classify the transitions myself.

```{r transitions}
# make the transitions df
trans <- df %>% 
  group_by(rna, id) %>% 
  mutate(
    from = state,
    to = lead(from), # get what the current state will transition to (next row)
    new_state = lag(from) != from, 
    new_state = if_else(is.na(new_state), FALSE, new_state),
    state_no = 1 + cumsum(new_state)
  ) %>% 
  group_by(rna, id, state_no) %>% 
  summarise(
    frames = n(),         # number of frames in this state
    time = n() / 5,       # numer of frames in seconds (frames = 0.2 s)
    from = last(from),    # get the last entry in from
    to = last(to)         # get the last entry in to
  )

trans
```

### Testing plotting of TDP

Replicate the transition-density plots with the code below.

```{r test-TDP}
trans %>% 
  # filter(rna == 1) %>% 
  filter(from > -0.2 & from < 1.2, 
         to > -0.2 & to < 1.2) %>% 
  ggplot(aes(from, to)) + 
  stat_density_2d(
    aes(fill = ..ndensity..,
        alpha = ..ndensity..),
    geom = "raster",
    n = 100,
    contour = FALSE, 
    h = c(0.2,0.2)
  ) +
  scale_fill_viridis_c() +
  scale_alpha_continuous(range = c(0.01, 1)) +
  
  coord_cartesian(xlim = c(0,1), ylim = c(0,1)) + 
  
  guides(alpha = "none", fill = "none") +
  
  facet_wrap(~rna) +
  
  
  theme_light() +
  theme(
    aspect.ratio = 1, 
    # panel.background = element_rect(fill = "gray90"), 
    panel.grid = element_line(colour = "gray70")
    )
```


# Fitting the decays

```{r}
cut_df <- trans %>% 
  # filter(rna == 1) %>% 
  mutate(
    from = case_when(
      from > 1 ~ 1, 
      from < 0 ~ 0, 
      TRUE ~ from
    ), 
    to = case_when(
      to > 1 ~ 1, 
      to < 0 ~ 0, 
      TRUE ~ to
    ),
    from_cut = cut(from, breaks = c(0.05, 0.49, 0.7, 1), labels = FALSE),
    to_cut   = cut(to, breaks = c(0.05, 0.49, 0.7, 1), labels = FALSE)
  )

cut_df %>% 
  filter(rna == 1) %>% 
  filter(!is.na(from_cut) & !is.na(to_cut)) %>% 
  ggplot(aes(time)) + 
  geom_histogram(aes(y = ..ncount..), bins = 50) + 
  facet_grid(
    rows = vars(from_cut), 
    cols = vars(to_cut)
  ) +
  xlim(c(0, 50))
```

```{r}
cut_df %>% 
  # filter(rna == 1) %>% 
  filter(from_cut == 2) %>%
  group_by(rna, id) %>% 
  summarise(time = sum(time)) %>% head
  ggplot(aes(y = time)) + 
  # geom_point() +
  # geom_density(aes(x =..ndensity..), n = 100, bw = 2) + 
  # geom_density(aes(x = -..ndensity..), n = 100, bw = 2) + 
  ggbeeswarm::geom_beeswarm(aes(x = 0), cex = 1.5) +
  facet_wrap(~rna) + 
  coord_cartesian(ylim  = c(0,50), xlim = c(-0.2, 0.2))
```

