---
title: "Turning the data into formatted data frames."
author: "Brady Johnston"
date: "2021-09-30"
output: 
    workflowr::wflow_html:
      code_folding: show
editor_options:
  chunk_output_type: console
---

```{r, echo=FALSE, message=FALSE}
library(tidyverse)

```


# smFRET Traces

The traces for the individual traces are in the following:
```{r}
paths <- c(
  "0" = "data/smfret-data/RNA0/rename_path",
  "1" = "data/smfret-data/RNA1_target",
  "2" = "data/smfret-data/RNA2",
  "3" = "data/smfret-data/RNA3",
  "4" = "data/smfret-data/RNA4",
  "5" = "data/smfret-data/RNA5",
  "6" = "data/smfret-data/RNA6",
  "7" = "data/smfret-data/RNA7",
  "8" = "data/smfret-data/RNA8",
  "9" = "data/smfret-data/RNA9",
  "10" = "data/smfret-data/RNA10",
  "11" = "data/smfret-data/RNA11",
  "12" = "data/smfret-data/RNA12",
  "13" = "data/smfret-data/RNA13"
)

head(paths)
```
```{r}
read_fun <- function(file) {
  some_list <- pbapply::pblapply(seq_along(file), function(x) {
    df <- readr::read_table(file[x], 
                            col_types = readr::cols(), 
                            col_names = c("frame", "don", "acc", "fret", "state"))
    df$mol <- x
    df
  })
  
  do.call(rbind, some_list)
}
# read the first set of frames as test
read_fun(list.files(paths[1], ".dat", full.names = TRUE))

```

## Read all of the Files in
```{r, eval=FALSE}
pbapply::pblapply(seq_along(paths), function(x) {
  df <- read_fun(list.files(paths[x], ".dat", full.names = TRUE))
  df$rna <- names(paths)[x]
  df
}) %>% 
  do.call(rbind, .) -> df

df

saveRDS(df, file = "data/all-traces.rds")

live_flow <- read_fun(list.files(here::here("data/smfret-data/RNA1_target/liveflow_RNA1/"), full.names = TRUE))

saveRDS(live_flow, file = "data/live-flow.rds")

```

## Calculate all of the state-changes
```{r}
traces <- readr::read_rds("data/all-traces.rds")

transitions <- traces %>% 
  group_by(rna, mol) %>% 
  mutate(
    from = state,
    to = lead(from), # get what the current state will transition to (next row)
    new_state = lag(from) != from, 
    new_state = if_else(is.na(new_state), FALSE, new_state),
    state_no = 1 + cumsum(new_state)
  ) %>% 
  group_by(rna, mol, state_no) %>% 
  summarise(
    frames = n(),         # number of frames in this state
    time = n() / 5,       # numer of frames in seconds (frames = 0.2 s)
    from = last(from),    # get the last entry in from
    to = last(to)         # get the last entry in to
  )

transitions

saveRDS(transitions, "data/transitions.rds")
```
## Cutting the states
```{r}
state_cutoffs <- c(0, 0.49, 0.7, 1)

cut_df <- transitions %>% 
  filter(time < 21) %>% 
  mutate(
    from = case_when(
      from > 1 ~ 1, 
      from < 0 ~ 0, 
      TRUE     ~ from
    ), 
    to = case_when(
      to > 1 ~ 1, 
      to < 0 ~ 0, 
      TRUE   ~ to
    ),
    from_cut = cut(from, breaks = state_cutoffs, labels = FALSE),
    to_cut   = cut(to, breaks = state_cutoffs, labels = FALSE)
  )

cut_df %>%
  filter(is.na(to))
```


## Bin the transitions from state 3
```{r}
state_3 <- cut_df %>% 
  filter(from_cut == 3) %>% 
  ungroup() %>% 
  mutate(
    time_bin = cut(time, breaks = seq(0, max(ceiling(time)), by = 1), labels = FALSE)
  ) %>% 
  group_by(rna, time_bin) %>% 
  summarise(n = n()) %>% 
  mutate(nn = n / max(n))

drc::drm(
  nn ~ time_bin, 
  data = filter(state_3, rna == 1), 
  fct = drc::EXD.2(names = c("max", "rate"))
)
```

## Fit exponential decay models to the binned times
```{r}
fitted_3 <- state_3 %>% 
  biochemr::b_dose_resp(time_bin, nn, rna, 
                        .model = drc::EXD.2(names = c("max", "rate"))) %>% 
  mutate(rna = as.numeric(rna))
```

## Plot the decays and the calculated half lives.
```{r}
decay_plot <- fitted_3 %>% 
  biochemr::b_plot() + 
  facet_wrap(~rna, ncol = 1, strip.position = "left") +
  # theme_void() + 
  geom_col(
    data = fitted_3 %>% unnest(data), 
    aes(dose, resp, fill = factor(rna)), 
    alpha = 0.3
  ) + 
  guides(fill = "none") +
  theme_classic() + 
  theme(
    axis.text.y = element_blank(), 
    axis.ticks.y = element_blank(), 
    strip.text = element_blank(), 
    strip.background = element_blank()
  )
# decay_plot

rate_plot <- fitted_3 %>% 
  biochemr::b_coefs() %>% 
  filter(term == "rate") %>% 
  ggplot(aes(estimate, 
             reorder(rna, -as.numeric(rna)), 
             colour = factor(rna))) + 
  geom_errorbarh(aes(
    xmin = estimate - std.error, 
    xmax = estimate + std.error
  ), 
   colour = "black", 
  height = 0.2) + 
  geom_point(size = 3) + 
  theme_light() + 
  coord_cartesian(xlim = c(0, 7)) +
  guides(colour = "none") + 
  # facet_wrap(~factor(rna), ncol = 1, scales = "free_y") + 
  theme(strip.background = element_blank(), 
        strip.text = element_blank()) + 
   labs(x = "Half Life (s)", 
        y = "RNA Sample")

patchwork::wrap_plots(rate_plot, decay_plot, ncol = 2, widths = c(2,1))
```

## Need to fix this error.

Currently the binning is discarding those states which didn't transition. For the molecules that make it to the higher state and the do no transition into lower states, despite staying there for a long period of time, this would be severely biasing the models generating to the shorter time periods that don't include these long-lived molecules in their states.

Need to re-bin them, and include those that don't transition but are just in the last state for a long time.
```{r}

cut1 <- 0.49
cut2 <- 0.7

cut_df %>% 
  filter(rna == 1) %>% 
  ggplot(aes(from, to)) + 
  geom_density2d_filled(
    contour_var = "ndensity",
    bins = 20,
    n = 80
    ) + 
  geom_segment(aes(x = x1, xend = x2, y = y1, yend = y2), 
               colour = "white", 
               linetype = "dashed",
               data = data.frame(
                 x1 = c(cut1, cut2, 0, 0, 0), 
                 x2 = c(cut1, cut2, cut1, cut2, 1),
                 y1 = c(0, 0, cut1, cut2, 0), 
                 y2 = c(cut1, cut2, cut1, cut2, 1)
               )) +
  scale_x_continuous(expand = expansion(c(0,0))) + 
  scale_y_continuous(expand = expansion(c(0,0))) + 
  theme_light() + 
  guides(fill = "none") +
  labs(x = "Before Transition", y = "After Transition") -> tdp_plot
```

```{r}
clusters <- cut_df %>%
  ungroup() %>% 
  filter(rna == 1) %>% 
  select(from, to) %>% 
  drop_na() %>% 
  kmeans(7, nstart = 100)
  

tdp_plot +
  geom_point(
    data = clusters$centers %>% 
  as.data.frame(), 
  mapping = aes(from, to), 
  size = 4, 
  colour = "white"
  )

traces %>% 
  filter(rna == 1) %>% 
  filter(fret < 1.2, fret > -0.2) %>%
  ggplot(aes(frame / 5, fret)) + 
  # geom_density2d_filled(
  #   aes(alpha = after_stat(density)),
  #   contour_var = "ndensity", 
  #   bins = 20, 
  #   n = 80
  # ) + 
  stat_density_2d(
    geom = "raster", 
    contour = FALSE, 
    contour_var = "ndensity", 
    mapping = aes(fill = ..ndensity.., 
                  alpha = ..ndensity..)
  ) +
  scale_alpha_continuous(range = c(0.2, 0.9)) + 
  scale_fill_viridis_c(option = "D", direction = 1) +
  stat_density_2d() +
  theme_light() +
  coord_cartesian(ylim = c(0,1))
```



# Fitting Gaussians with `{mixtools}` to the histograms

```{r}
library(mixtools)

get_gaussians <- function(x, mu = c(0.3, 0.6, 0.9), sigma = 0.05) {
  normalmixEM(x, mu = mu, sigma = sigma, maxit = 1e4, maxrestarts = 40) %>% 
    .[c("mu", "sigma", "lambda")] %>% 
    as.data.frame()
}

add_gaussians <- function(plot, gaussians) {
  purrr::reduce(seq(nrow(gaussians)),
                ~ .x + geom_function(
                  fun = dnorm,
                  args = list(mean = gaussians[.y, "mu"],
                              sd = gaussians[.y, "sigma"]),
                  mapping = aes(
                    y = after_stat(y) / max(after_stat(y)) * gaussians[.y, "lambda"] / max(gaussians[, "lambda"]),
                    colour = as.character(round(gaussians[.y, "mu"], 2))
                  )
                  
                ),
                .init = plot + labs(colour = "Peak", y = "Density"))
}


gauss_plot <- function(x, mu = c(0.3, 0.6, 0.9), sigma = 0.05) {
  gauss <- get_gaussians(x, mu = mu, sigma = sigma)
  
  ggplot(data.frame(x = x), aes(x = x)) + 
    geom_histogram(colour = "gray20",
                   fill = "gray90",
                   binwidth = 0.05,
                   aes(y = ..ncount..),
                   alpha = 0.4) + 
    theme(legend.position = c(0.9,0.9), 
          legend.justification = c(1,1)) +
    theme_light() -> plt
  
  plt %>% 
    add_gaussians(gauss)
}



binned <- traces %>% 
  filter(fret < 1, fret > 0) %>% 
  group_by(rna)




# future::plan(future::multisession, workers = 7)
# tictoc::tic()
# progressr::with_progress({
#   p <- progressr::progressor(14)
#   
#   plotted <- binned %>% 
#     group_by(rna) %>% 
#     nest() %>% 
#     ungroup() %>% 
#     mutate(
#       plot = furrr::future_map(data, function(x) {
#         p()
#         gauss_plot(pull(x, fret)) %>% print
#       } 
#     ))
# })
# tictoc::toc()
```

What about fitting gaussians to less observations, say we bin it first then we fit.

```{r}
plt12 <- traces %>%
  filter(rna == 12, fret < 1, fret > 0) %>% 
  mutate(
    binned = cut(fret, breaks = seq(0, 1, by = 0.05), labels = FALSE) * 0.05
  ) %>% 
  group_by(binned) %>% 
  # summarise(n = n()) %>% 
  pull(binned) %>% 
  gauss_plot(mu = c(0.2, 0.4, 0.6, 0.8)) + 
  theme(legend.position = c(0.1,0.9),
        legend.justification = c(0, 1))
plt12 + 
  theme(legend.position = c(0.1,0.9),
        legend.justification = c(0, 1))

plt0 <- traces %>%
  filter(rna == 0, fret < 1, fret > 0) %>% 
  mutate(
    binned = cut(fret, breaks = seq(0, 1, by = 0.05), labels = FALSE) * 0.05
  ) %>% 
  group_by(binned) %>% 
  # summarise(n = n()) %>% 
  pull(binned) %>% 
  gauss_plot(mu = c(0.2, 0.4, 0.6, 0.8)) + 
  theme(legend.position = c(0,1),
        legend.justification = c(-0.1, 1.1), 
        legend.background = element_rect(colour = "gray20", 
                                         fill = "white"))

patchwork::wrap_plots(plt0, plt12)

```

