---
title: "Biophysical Studies of dsnPPR10"
author: "Brady Johnston"
date: "2021-10-06"
output: 
    workflowr::wflow_html:
      code_folding: hide
editor_options:
  chunk_output_type: console
---

Setup code.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
knitr::opts_chunk$set(warning = FALSE, 
                      autodep = TRUE
                      )

traces <- readRDS("data/all-traces.rds")
transitions <- readRDS("data/transitions.rds")
spr <- readRDS("~/Dropbox/BondLab/Data/SPR/spr.rds")
as_tibble() %>% 
  mutate(rna = str_trim(rna)) %>% 
  mutate(
    rna = factor(rna,
                 levels = reorder(rna, str_extract(rna, "\\d+")), 
                 labels = paste("RNA", str_pad(1:13, 2, "left", pad = "0"), sep = ""))
  )
spr_coef <- read.csv("~/Dropbox/BondLab/Data/SPR/fitted-data.csv") %>% 
  as_tibble() %>% 
  mutate(rna = str_trim(rna)) %>% 
  mutate(
    rna = factor(rna,
                 levels = reorder(rna, str_extract(rna, "\\d+")), 
                 labels = paste("RNA", str_pad(1:13, 2, "left", pad = "0"), sep = ""))
  )
live_flow <- readRDS("data/live-flow.rds")
```

#### Useful Functions

```{r functions}
time_density_plot <- function(data) {
  data %>% 
    filter(fret > 0, fret < 1) %>% 
    mutate(frame = frame / 5) %>% 
    ggplot(aes(frame, fret)) + 
    geom_point(alpha = 0.1, colour = "gray30") +
    stat_density_2d(aes(fill = ..ndensity.., alpha = ..ndensity..),
                    geom = "raster",
                    n = 200,
                    contour = FALSE) +
    geom_density_2d(bins = 10, colour = "white", contour_var = "ndensity") +
    scale_fill_viridis_c(option = "D",
                         direction = 1,
                         limits = c(0.002, NA),
                         # trans = "log10",
                         na.value = NA) +
    scale_alpha_continuous(range = c(0.01, 0.8)) +
    coord_cartesian(ylim = c(0,1), xlim = c(0,100)) + 
    guides(alpha = "none", 
           fill = "none") +
    # theme_light() +
    theme_linedraw() +
    labs(x = "Time (s)", y = "FRET") +
    scale_x_continuous(expand = c(0,0)) + 
    scale_y_continuous(expand = c(0,0))
}
tdp_plot <- function(data) {
  data %>% 
    filter(from > -0.2 & from < 1.2, 
         to > -0.2 & to < 1.2) %>% 
  ggplot(aes(from, to)) + 
  stat_density_2d(
    aes(fill = ..ndensity..,
        alpha = log2(..ndensity.. + 1)),
    geom = "raster",
    n = 400,
    contour = FALSE,
    h = c(0.2,0.2)
  ) +
  geom_density2d(
    contour_var = "ndensity",
    n = 100,
    bins = 15,
    colour = "black", alpha = 0.3
  ) +
  scale_fill_viridis_c(option = "B", direction = 1) +
  scale_alpha_continuous(range = c(0, 1)) +
  
  coord_cartesian(xlim = c(0,1), ylim = c(0,1)) + 
  
  guides(alpha = "none", fill = "none") +
  
  scale_x_continuous(expand = expansion(0), breaks = seq(0,1,0.2)) +
  scale_y_continuous(expand = expansion(0), breaks = seq(0,1,0.2)) +
  
  labs(x = "Before Transition", 
       y = "After Transition", 
       y = "Transition Density Plot") +
  
  theme_light() +
  theme(
    aspect.ratio = 1, 
    # panel.background = element_rect(fill = "gray90"), 
    panel.grid = element_line(colour = "gray90")
    )
}

get_gaussians <- function(x,
                          mu = c(0.3, 0.6, 0.9),
                          sigma = 0.05) {
  mixtools::normalmixEM(
    x,
    mu = mu,
    sigma = sigma,
    maxit = 1e4,
    maxrestarts = 40
  ) %>%
    .[c("mu", "sigma", "lambda")] %>%
    as.data.frame()
}

add_gaussians <- function(plot, gaussians) {
  purrr::reduce(seq(nrow(gaussians)),
                ~ .x + geom_function(
                  fun = dnorm,
                  args = list(mean = gaussians[.y, "mu"],
                              sd = gaussians[.y, "sigma"]),
                  mapping = aes(
                    y = after_stat(y) / max(after_stat(y)) * gaussians[.y, "lambda"] / max(gaussians[, "lambda"]),
                    colour = as.character(round(gaussians[.y, "mu"], 2))
                  )
                  
                ),
                .init = plot + labs(colour = "Peak", y = "Density"))
}


gauss_plot <- function(x, mu = c(0.3, 0.6, 0.9), sigma = 0.05) {
  gauss <- get_gaussians(x, mu = mu, sigma = sigma)
  
  ggplot(data.frame(x = x), aes(x = x)) + 
    geom_histogram(colour = "gray20",
                   fill = "gray90",
                   binwidth = 0.05,
                   aes(y = ..ncount..),
                   alpha = 0.4) + 
    theme(legend.position = c(0.9,0.9), 
          legend.justification = c(1,1)) +
    theme_light() -> plt
  
  plt %>% 
    add_gaussians(gauss)
}

```

# smFRET of Target

## TDP

Single-molecule FRET was carried out on dsnPPR10 with multiple RNA constructs.

```{r}
#| fig.cap: Transition density plot for the target RNA sample from smFRET experiments.
transitions %>% 
  filter(rna == 1) %>% 
  tdp_plot() + 
  labs(title = "Transition Density Plot")

```

From the transition density plot, there are appears to be three distinct states that the molecule visits when the target RNA is in the solution. There is a *low* *FRET* below \~ 0.5, there is a *mid-FRET* state between \~0.5 and \~ 0.7 and there is a *high-FRET* state that exists \> 0.7 FRET.

This suggests that there may be an intermediate in the binding pathway of the protein, going from the unbound "open" *low-FRET* state to the medium, and then onto the *high-FRET* state as the final state of binding where the majority of the molecules spend their time.

## FRET Histogram

The FRET histogram for these molecules looks like below:

```{r rna01-fret-histogram, warning=FALSE}
#| fig.cap: FRET histogram of the protein with its RNA target.

traces %>% 
  filter(rna == 1) %>% 
  ggplot(aes(fret)) + 
  xlim(c(0,1)) + 
  geom_histogram(
    aes(y = ..ncount..),
    binwidth = 0.05,
    colour = "gray20", 
    fill = "gray90") + 
  theme_light() + 
  labs(x = "FRET", y = "Relative Count")
```

We can fit three gaussians to this FRET distribution, to see how the three potential states might be contributing to the overall distribution.

```{r rna01-fret-histogram-gauss}
#| fig.cap: FRET histogram of the protein with it's RNA target, with three gaussian distributions fitted fo the data.
traces %>% 
  filter(rna == 1, fret <= 1, fret >= 0) %>% 
  pull(fret) %>% 
  gauss_plot(mu = c(0.3, 0.6, 0.9)) + 
  theme(legend.position = c(0.1,0.9), 
        legend.justification = c(0,1), 
        legend.background = element_rect(colour = "gray50"))
```

## Time-Density Plot

This same pattern of three distinct states is also apparent in the time-resolved FRET density plots, showing the density of FRET of all of the molecules observed over time.

```{r time-density-rna01, warning=FALSE}
#| fig.cap: FRET distribution over time after target RNA is injected at the 5 s mark. Plot shows a distinct mid-to-high FRET state that most of the molecules visit before moving to the highest FRET state.

live_flow %>% 
  filter(fret > -0.2, fret < 1.2) %>% 
  time_density_plot()
```

# SPR of Target

Along with the smFRET experiments, data from Surface Plasmon Resonance experiments were also collected. These experiments showed binding and association of the protein with all 13 of the different RNA, show distinct association and dissociation of the protein with the different targets depending on their differences to the target sequence.

```{r spr-data}
spr[seq(1, nrow(spr), 10), ] %>% 
  ggplot(aes(time, Response, colour = RNA)) +
  geom_line() + 
  theme_light() + 
  theme(legend.position = "bottom") + 
  labs(x = "Time (s)", 
       y = "Response Units (RU)")
```

The target RNA (RNA01) shows the best binding and the slowest dissociation, as you would expect.

Interestingly, despite having a high binding affinity (K$_{d}$ = \~ 1 nM), the reaction does not reach equilibrium even after 300 seconds and the response units do not plateau. In the dissociation phase, it does not follow a typical single-exponential decay as you would expect from a single-step dissociation. However you can fit a double-exponential decay and get a reasonable fit to the dissociation curve, indicating that there is likely a two-step dissociation going on.

```{r}
df <- spr[seq(1, nrow(spr), 10), ] %>% 
  rename(rna = RNA, resp = Response) %>% 
  filter(rna == "RNA01", time > 300)

model <-  df %>% 
  nls(resp ~ SSasymp(time, R0 = A1, lrc = t1,  Asym = Asym), data = .)

model2 <- df %>% 
  nls(
    resp ~ SSbiexp(time, A1, t1, A2, t2), 
    data = .
  )

lines <- data.frame(
  time = df$time,
  "Single" = model$m$predict(), 
  "Double" = model2$m$predict()
) %>% 
  pivot_longer(c(Single, Double), values_to = "resp")

df %>% 
  ggplot(aes(time, resp)) + 
  geom_line(colour = "red", alpha = 0.5, size = 1) +
  geom_line(
    data = lines, 
    aes(linetype = name), 
    size = 1, 
    alpha = 0.6
  ) +
  scale_linetype_manual(values = c("dashed", "dotted")) +
  theme_light() + 
  labs(linetype = "Exponential Decay\nModel", 
       x = "Time (s)", 
       y = "Response Units (RU)") + 
  theme(legend.position = c(0.8, 0.8), 
        legend.background = element_rect(colour = "gray70"))
```

All of these results point to some intermediate step in the binding and unbinding process, that so far has not been captured by the bound and unbound structures of the dsnPPR10 protein, and no mention of such binding mechanisms or intermediates appears in the PPR literature.

```{r binding-diagram, fig.height=5, fig.align='center'}
#| fig.cap: Reaction diagram of proposed two-step binding interaction that PPR proteins undergo when interacting with RNA.
library(nomnoml)
nomnoml("#direction: down
[P + R]->Association [PR]
[PR]->Dissociation [P + R]
[PR]->PPR Tightening [PR*]
[PR*] -> PPR Loosening[PR]
")

```

The states in this diagram can be representative of the different FRET states that are observed.

```{r fret-diagram, fig.height=5, fig.align='center'}
#| fig.cap: Reaction diagram of proposed two-step binding interaction that PPR proteins undergo when interacting with RNA, with the states labelled by their respective FRET states.
nomnoml("#direction: down
[low-FRET]->Association [mid-FRET]
[mid-FRET]->Dissociation [low-FRET]
[mid-FRET]->PPR Loosening [high-FRET]
[high-FRET] -> PPR Tightening[mid-FRET]
")
```

## Rate Constants
### Dissociation Rates

From the fitting fo the exponential decay models, we are able to get rate constants for the loosening and the dissociation steps that the protein undergoes.

```{r half-life-plot, message=FALSE, error=FALSE, warning=FALSE}
#| fig.height: 7
#| fig.width: 8
#| fig.cap: Fitted rates of decay for the double-exponential model for each of RNA samples where there was a sufficient enough response to model. RNA2 and RNA5 did not show enough of a response to model.

some_theme <- function() {
  theme(
    panel.grid.major.y = element_blank(), 
    axis.title.y = element_blank(),
    axis.text.y = ggtext::element_markdown(family = "mono")
  )
}

t1plot <- spr_coef %>%
   
  # ggplot(aes(y = factor(rna, levels = reorder(rna, str_extract(rna, "\\d+"))))) +
  ggplot(aes(y = rna)) + 
  # geom_errorbarh(aes(xmin = t1 - t1.err, xmax = t1 + t1.err),
  #                 height = 0.4) +
  geom_point(aes(x = t1, colour = rna), size = 3) +
  ggtext::geom_richtext(
    data = mdfret::rna_table[1:13, ], 
    aes(label = html_seq, 
        y = paste("RNA", str_pad(1:13, 2, "left", pad = "0"), sep = ""),
        x = 0), 
    fill = NA,
    hjust = 0, 
    label.fill = NA, 
    label.color = NA,
    family = "mono", 
    size = 2.2, 
    nudge_y = 0.2
  ) + 
  theme_light() +
  labs(x = "t1 (s)", y = "RNA") + 
  guides(colour = "none") + 
  xlim(c(0,NA)) + 
  # scale_y_discrete(
  #   labels = mdfret::rna_table$html_seq[1:13]
  # ) + 
  some_theme()

t2plot <- spr_coef %>%
  ggplot(aes(y = factor(rna, levels = reorder(
    rna, str_extract(rna, "\\d+")
)))) +
  geom_errorbarh(aes(xmin = t2 - t2.err, xmax = t2 + t1.err),
                  height = 0.4) +
  geom_point(aes(x = t2, colour = rna), size = 3) +
  theme_light() +
  labs(x = "t2 (s)", y = "RNA") + 
  guides(colour = "none") + 
  xlim(c(0,NA)) +
  some_theme() +
  theme(axis.text.y = element_blank(), 
        axis.title.y = element_blank(), 
        axis.ticks.y = element_blank())

patchwork::wrap_plots(
  t1plot, 
  t2plot, 
  t2plot + labs(x = "log10(t2) (s)") + scale_x_log10()
  ) + patchwork::plot_annotation(
  title = "Rates of Decay", 
  subtitle = "Calcultaed half lives of for the double-exponential decay models."
)
```

The plotting of these values show some interesting groupings of the RNA that were tested. The `t1` values do not vary much between samples, indicating this is likely the rate change of the protein completely dissociating from the bound RNA. The three samples that do have noticeably longer half lives are RNA9-11, which are the the three RNA which are double the length of the other RNA targets.


### Association Rates

If we also look at the on-rates, we can see a story developing.
```{r}
spr_coef %>% 
  ggplot(aes(y = rna, x = kon)) + 
  geom_segment(aes(y = rna, yend = rna, x = 0, xend = kon), 
               colour = "black", 
               alpha = 0.3) +
  geom_point(aes(colour = rna), size = 3) + 
  # ggtext::geom_richtext(
  #   aes(label = html_seq, x = 0),
  #   data = mdfret::rna_table[1:13, ],
  #   nudge_y = 0.3,
  #   label.padding = unit(0, "lines"),
  #   hjust = 0, 
  #   fill = NA, 
  #   label.color = NA, 
  #   family = "mono"
  # ) + 
  labs(x = latex2exp::TeX("K_{on} M^{-1} s^{-1}")) +
  theme_light() + 
  some_theme() + 
  theme(legend.position = "")
```

### Koff
```{r}
koff_log10 <- spr_coef %>% 
  ggplot(aes(y = rna, x = koff)) + 
  geom_segment(aes(y = rna, 
                   x = 0, 
                   yend = rna, 
                   xend = koff), 
               colour = "black", 
               alpha = 0.3) +
  geom_point(aes(colour = rna), size = 3) + 
  labs(x = latex2exp::TeX("log10(k_{off}) M^{-1} s^{-1}")) +
  theme_light() + 
  some_theme() + 
  theme(legend.position = "") + 
  scale_x_log10(limits = c(100, NA)) + 
  simpleThemes::strip_axis("y")

koff <- spr_coef %>% 
  ggplot(aes(y = rna, x = koff)) + 
  geom_segment(aes(y = rna, 
                   x = 0, 
                   yend = rna, 
                   xend = koff), 
               colour = "black", 
               alpha = 0.3) +
  geom_point(aes(colour = rna), size = 3) + 
  # ggtext::geom_richtext(
  #   aes(label = html_seq, x = 0),
  #   data = mdfret::rna_table[1:13, ],
  #   nudge_y = 0.3,
  #   label.padding = unit(0, "lines"),
  #   hjust = 0, 
  #   fill = NA, 
  #   label.color = NA, 
  #   family = "mono"
  # ) + 
  labs(x = latex2exp::TeX("k_{off} M^{-1} s^{-1}")) +
  theme_light() + 
  some_theme() + 
  theme(legend.position = "")

patchwork::wrap_plots(koff, koff_log10)
```


## Fitting to the Association Curves

```{r}
spr[1:nrow(spr) %% 10 == 0, ] %>% 
  rename(resp = Response, rna = RNA) %>% 
  filter(time > 0, time < 300) -> values

  ggplot(aes(time, resp, colour = rna)) + 
  scale_colour_discrete(breaks = unique(values))
  geom_line()
```


```{r}
spr[1:nrow(spr) %% 10 == 0, ] %>% 
  rename(resp = Response, rna = RNA) %>% 
  filter(time > 0, time < 300) %>% 
  filter(rna == "RNA01") -> data1
  
  # plot
  # drc::drm(resp ~ time, fct = drc::EXD.2(), data = .) %>%
  # 
  data1 %>% 
  nls(resp ~ SSasymp(time, Asym = Asym, R0, lrc), data = .) -> model
model$m
par(col = "black")
plot(data1$time, data1$resp, col = "gray70")
lines(model$m$fitted(), col = "red")
```


```{r}
live_flow %>% 
  mutate(rna = "1") %>% 
  group_by(rna, mol) %>% 
  mutate(
    from = state,
    to = lead(from), # get what the current state will transition to (next row)
    new_state = lag(from) != from, 
    new_state = if_else(is.na(new_state), FALSE, new_state),
    state_no = 1 + cumsum(new_state)
  ) %>% 
  group_by(rna, mol, state_no) %>% 
  summarise(
    frames = n(),         # number of frames in this state
    time = n() / 5,       # numer of frames in seconds (frames = 0.2 s)
    from = last(from),    # get the last entry in from
    to = last(to)         # get the last entry in to
  ) %>% 
  tdp_plot() +
  coord_cartesian(xlim = c(0,1.2), ylim = c(0,1.2))
  # geom_point()
  # head
  ggplot(aes(from, to)) + 
  geom_point() +
  geom_density_2d(contour_var = "ndensity", bins = 20) + 
  xlim(c(0,1.4)) +
  ylim(c(0,1.4))
```

