---
title: "Biophysical Studies of dsnPPR10"
author: "Brady Johnston"
date: "2021-10-06"
output:
  workflowr::wflow_html:
  code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)


traces <- readRDS("data/all-traces.rds")
transitions <- readRDS("data/transitions.rds")
spr <- readRDS("~/Dropbox/BondLab/Data/SPR/spr.rds")
live_flow <- readRDS("data/live-flow.rds")
```

#### Useful Functions

```{r functions}
time_density_plot <- function(data) {
  data %>% 
    filter(fret > 0, fret < 1) %>% 
    mutate(frame = frame / 5) %>% 
    ggplot(aes(frame, fret)) + 
    geom_point(alpha = 0.1, colour = "gray30") +
    stat_density_2d(aes(fill = ..ndensity.., alpha = ..ndensity..),
                    geom = "raster",
                    n = 200,
                    contour = FALSE) +
    geom_density_2d(bins = 10, colour = "white", contour_var = "ndensity") +
    scale_fill_viridis_c(option = "D",
                         direction = 1,
                         limits = c(0.002, NA),
                         # trans = "log10",
                         na.value = NA) +
    scale_alpha_continuous(range = c(0.01, 0.8)) +
    coord_cartesian(ylim = c(0,1), xlim = c(0,100)) + 
    guides(alpha = "none", 
           fill = "none") +
    # theme_light() +
    theme_linedraw() +
    labs(x = "Time (s)", y = "FRET") +
    scale_x_continuous(expand = c(0,0)) + 
    scale_y_continuous(expand = c(0,0))
}
tdp_plot <- function(data) {
  data %>% 
    filter(from > -0.2 & from < 1.2, 
         to > -0.2 & to < 1.2) %>% 
  ggplot(aes(from, to)) + 
  stat_density_2d(
    aes(fill = ..ndensity..,
        alpha = log2(..ndensity.. + 1)),
    geom = "raster",
    n = 400,
    contour = FALSE,
    h = c(0.2,0.2)
  ) +
  geom_density2d(
    contour_var = "ndensity",
    n = 100,
    bins = 15,
    colour = "black", alpha = 0.3
  ) +
  scale_fill_viridis_c(option = "B", direction = 1) +
  scale_alpha_continuous(range = c(0, 1)) +
  
  coord_cartesian(xlim = c(0,1), ylim = c(0,1)) + 
  
  guides(alpha = "none", fill = "none") +
  
  scale_x_continuous(expand = expansion(0), breaks = seq(0,1,0.2)) +
  scale_y_continuous(expand = expansion(0), breaks = seq(0,1,0.2)) +
  
  labs(x = "Before Transition", 
       y = "After Transition", 
       y = "Transition Density Plot") +
  
  theme_light() +
  theme(
    aspect.ratio = 1, 
    # panel.background = element_rect(fill = "gray90"), 
    panel.grid = element_line(colour = "gray90")
    )
}

get_gaussians <- function(x,
                          mu = c(0.3, 0.6, 0.9),
                          sigma = 0.05) {
  mixtools::normalmixEM(
    x,
    mu = mu,
    sigma = sigma,
    maxit = 1e4,
    maxrestarts = 40
  ) %>%
    .[c("mu", "sigma", "lambda")] %>%
    as.data.frame()
}

add_gaussians <- function(plot, gaussians) {
  purrr::reduce(seq(nrow(gaussians)),
                ~ .x + geom_function(
                  fun = dnorm,
                  args = list(mean = gaussians[.y, "mu"],
                              sd = gaussians[.y, "sigma"]),
                  mapping = aes(
                    y = after_stat(y) / max(after_stat(y)) * gaussians[.y, "lambda"] / max(gaussians[, "lambda"]),
                    colour = as.character(round(gaussians[.y, "mu"], 2))
                  )
                  
                ),
                .init = plot + labs(colour = "Peak", y = "Density"))
}


gauss_plot <- function(x, mu = c(0.3, 0.6, 0.9), sigma = 0.05) {
  gauss <- get_gaussians(x, mu = mu, sigma = sigma)
  
  ggplot(data.frame(x = x), aes(x = x)) + 
    geom_histogram(colour = "gray20",
                   fill = "gray90",
                   binwidth = 0.05,
                   aes(y = ..ncount..),
                   alpha = 0.4) + 
    theme(legend.position = c(0.9,0.9), 
          legend.justification = c(1,1)) +
    theme_light() -> plt
  
  plt %>% 
    add_gaussians(gauss)
}

```

# The Target RNA

Single-molecule FRET was carried out on dsnPPR10 with multiple RNA constructs.

```{r}
#| fig.cap: Transition density plot for the target RNA sample from smFRET experiments.
transitions %>% 
  filter(rna == 1) %>% 
  tdp_plot() + 
  labs(title = "Transition Density Plot")

```

From the transition density plot, there are appears to be three distinct states that the molecule visits when the target RNA is in the solution. There is a *low* *FRET* below \~ 0.5, there is a *mid-FRET* state between \~0.5 and \~ 0.7 and there is a *high-FRET* state that exists \> 0.7 FRET.

This suggests that there may be an intermediate in the binding pathway of the protein, going from the unbound "open" *low-FRET* state to the medium, and then onto the *high-FRET* state as the final state of binding where the majority of the molecules spend their time.

The FRET histogram for these molecules looks like below:

```{r rna01-fret-histogram}
#| fig.cap: FRET histogram of the protein with its RNA target.

traces %>% 
  filter(rna == 1) %>% 
  ggplot(aes(fret)) + 
  xlim(c(0,1)) + 
  geom_histogram(
    aes(y = ..ncount..),
    colour = "gray20", 
    fill = "gray90") + 
  theme_light() + 
  labs(x = "FRET", y = "Relative Count")
```

We can fit three gaussians to this FRET distribution, to see how the three potential states might be contributing to the overall distribution.

```{r rna01-fret-histogram-gauss}
#| fig.cap: FRET histogram of the protein with it's RNA target, with three gaussian distributions fitted fo the data.
traces %>% 
  filter(rna == 1, fret <= 1, fret >= 0) %>% 
  pull(fret) %>% 
  gauss_plot(mu = c(0.3, 0.6, 0.9)) + 
  theme(legend.position = c(0.1,0.9), 
        legend.justification = c(0,1), 
        legend.background = element_rect(colour = "gray50"))
```

This same pattern of three distinct states is also apparent in the time-resolved FRET density plots, showing the density of FRET of all of the molecules observed over time.

```{r time-density-rna01}
#| fig.cap: FRET distribution over time after target RNA is injected at the 5 s mark. Plot shows a distinct mid-to-high FRET state that most of the molecules visit before moving to the highest FRET state.

live_flow %>% 
  filter(fret > -0.2, fret < 1.2) %>% 
  time_density_plot()
```
